---
title: "Gene Level Comparison"
author: "Rebecca L"
date: "2025-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggpubr)
library(ggpattern)
library(tidyverse)
library(ggplot2)
library(pheatmap)
library(DESeq2)
library(kableExtra)
library(vsn)
library(ggfortify)
library(AnnotationDbi)
library (org.Hs.eg.db)
library(PCAtools)
library(reshape2)
library(RColorBrewer)
library(fgsea)
library(GOSemSim)
library(enrichplot)
library(clusterProfiler)
library(ggvenn)
library(gplots)
library(purrr)
library(biomaRt)
library(GSEABase)
library(GSVA)
source("customfunc.R")

```
```{r import, normalisation, PCA plot}

#Count matrix
cdna_count=read.table("cdna2_SKMM2_counts_gene.txt", header=T, sep="\t")[1:7]
drna_count=read.table("drnasub_SKMM2_counts_gene.txt", header=T, sep="\t")
combdata=left_join(cdna_count, drna_count, by="GENEID")

#Annotate HGNC symbols
gene_info=read.csv("tx_v46_info.csv", header=TRUE)[,c(2:3)]
readcounts_an=left_join(combdata, gene_info, by=join_by(GENEID==GENENAME), multiple="any")
rownames(readcounts_an)=readcounts_an$GENEID
#Sample Information
sampleinfo=read.table("sampleinfo.csv", header=T, row.names=1, sep=",", stringsAsFactors = TRUE)
#Lib size plot
librarySizes <- colSums(readcounts_an[,2:13])
barplot(librarySizes, names=names(librarySizes), las=2, main="Barplot of library sizes")

#Filter out 'IG' genes
ig=grep('^IGH|^IGK|^IGL|^HLA', readcounts_an$HGNC)
readcounts_an=readcounts_an[-ig,]

#Make DESeq Dataset & Filter No Counts
readcounts_an[,2:13]=sapply(readcounts_an[,2:13],as.integer)
dds=DESeqDataSetFromMatrix(countData=readcounts_an[,2:13], 
                           colData=sampleinfo, 
                           design=~Experiment+Condition)

hgnc=readcounts_an$HGNC
mcols(dds)= DataFrame(mcols(dds), hgnc)
mcols(dds)

#DESeq Object
dds <- estimateSizeFactors(dds)
dds=dds[rowSums(counts(dds))>=1,]

#DESeq + VSD Normalisation
dds=DESeq(dds)
vsd=vst(dds, blind=TRUE)
vsd_mat=as.matrix(assay(vsd))
meanSdPlot(vsd_mat, ranks=FALSE)


#PCA Plot
pca_res=prcomp(t(vsd_mat), scale. = TRUE)
autoplot(pca_res, data=sampleinfo, colour="Condition", shape="Experiment", size=4)



sampleDists <- dist(t(assay(vsd)))
library("RColorBrewer")
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$Experiment, vsd$Condition, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Reds")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)

```
```{r import DGE results}
#Read in results
cdna_dge=read.csv("DGE_cdna.csv", row.names=1, header=TRUE)
drna_dge=read.csv("DGE_drnasub.csv", row.names=1, header=TRUE)

#Remove NA
cdna_dge=cdna_dge[!is.na(cdna_dge$padj),]
drna_dge=drna_dge[!is.na(drna_dge$padj),]

#Extract sig up and sig down genes
cdna_up=rownames(cdna_dge)[cdna_dge$padj<0.05 & cdna_dge$log2FoldChange>0]
drna_up=rownames(drna_dge)[drna_dge$padj<0.05 & drna_dge$log2FoldChange>0]
up=list("PCR-cDNA"=cdna_up, directRNA=drna_up)
rm(cdna_up, drna_up)

cdna_down=rownames(cdna_dge)[cdna_dge$padj<0.05 & cdna_dge$log2FoldChange<0]
drna_down=rownames(drna_dge)[drna_dge$padj<0.05 & drna_dge$log2FoldChange<0]
down=list("PCR-cDNA"=cdna_down, directRNA=drna_down)
rm(cdna_down, drna_down)

#Venns
library(ggvenn)

#UP
ggvenn(up, 
       fill_color=c("turquoise3", "navy"),
       text_size=6,
       show_percentage=FALSE, 
       auto_scale=TRUE)
ggvenn(down, 
       fill_color=c("turquoise3", "navy"),
       text_size=6,
       show_percentage=FALSE, 
       auto_scale=TRUE) 

```

```{r gsva}

#Get entrez IDs
ensembl=useEnsembl(biomart="genes",
                   dataset="hsapiens_gene_ensembl")

G_list=getBM(attributes=c("ensembl_gene_id_version", "entrezgene_id", "hgnc_symbol"), values=readcounts_an$GENEID, mart=ensembl)
gsva_array=as.data.frame(vsd_mat)
gsva_array$geneid=rownames(gsva_array)
gsva_an=inner_join(gsva_array, G_list, by=join_by(geneid==ensembl_gene_id_version), multiple="any")
gsva_an=gsva_an[!duplicated(gsva_an$hgnc_symbol),]
rownames(gsva_an)=gsva_an$hgnc_symbol


#Set up array
pdata=AnnotatedDataFrame(sampleinfo)
exprs=ExpressionSet(assayData=as.matrix(gsva_an[,1:12]), phenoData=pdata)


#Hallmark gene set
hallmark=list.files()
h.gs=getGmt("h.all.v2024.1.Hs.symbols.gmt", collectionType=BroadCollection(category="h"),
            geneIdType = SymbolIdentifier())

gsva_par=gsvaParam(exprs, h.gs, minSize=5, maxSize=500)
gsva=gsva(gsva_par)

#DE analysis of gene sets
library(limma)
library(edgeR)
gsva$group=paste0(gsva$Experiment,gsva$Condition)
mod=model.matrix(~gsva$group)
contrasts=makeContrasts("cDNAControl-cDNAIL6", "dRNAControl-dRNAIL6", levels=c("cDNAIL6", "cDNAControl", "dRNAIL6", "dRNAControl"))

fit=lmFit(gsva, mod)
fit=contrasts.fit(fit, contrasts)
fit=eBayes(fit)

res=decideTests(fit, p.value=0.05)
summary(res)
tt=topTable(fit, coef=1, n=50)
tt2=topTable(fit, coef=2, n=50)

#Format results df
pathways=rownames(tt)
pathways=strsplit(pathways, split="HALLMARK_" )
rownames(tt)=lapply(pathways, '[', 2)
tt=rownames_to_column(tt, var="pathway")
tt$cdna_sig=case_when(tt$adj.P.Val<0.05 & tt$logFC>0 ~'UP', tt$adj.P.Val<0.05 & tt$logFC<0 ~"DOWN", .default="NS")
tt=tt[,c(1,8)]

pathways=rownames(tt2)
pathways=strsplit(pathways, split="HALLMARK_")
rownames(tt2)=lapply(pathways, "[", 2)
tt2=rownames_to_column(tt2, var="pathway")
tt2$drna_sig=case_when(tt2$adj.P.Val<0.05 & tt2$logFC>0 ~'UP', tt2$adj.P.Val<0.05 & tt2$logFC<0 ~"DOWN", .default="NS")
tt2=tt2[, c(1,8)]

ann=inner_join(tt, tt2, by="pathway")
ann=column_to_rownames(ann, var="pathway")
#Annotate each gene set with up down or NS

ann_colour=list(cdna_sig=c(NS="grey30", UP="red", DOWN="mediumblue"),
                drna_sig=c(NS="grey30", UP="red", DOWN="mediumblue"),
                Condition=c(Control="pink", IL6="darkred"),
                Experiment=c(cDNA="turquoise3", dRNA="navy"))




pathways=rownames(gsva)
pathways=strsplit(pathways, split="HALLMARK_" )
gsva_array=exprs(gsva)
rownames(gsva_array)=lapply(pathways, '[', 2)

heatmap=pheatmap(gsva_array, clustering_distance_cols="correlation", clustering_distance_rows="correlation", annotation_colors=ann_colour, annotation_row=ann, annotation_col=sampleinfo, cutree_cols=2, cutree_rows=2, scale="column")
```


```{r scatterplots illustrative}

#Annotate protocol and bind
dge_comb=left_join(cdna_dge[,c(2,7)], drna_dge[,c(2,7)],by="hgnc")
colnames(dge_comb)=c("PCR-cDNA", "hgnc", "DirectRNA")
dge_comb_m=pivot_longer(dge_comb, cols=c(1,3))

ggplot(dge_comb, aes(x=`PCR-cDNA`, y=DirectRNA))+geom_point(colour="grey30")+
  scale_y_continuous(limits=c(-2.5,5))+
  scale_x_continuous(limits=c(-2.5,5))+theme(
      legend.text = element_text(size =14),
      axis.text = element_text(size = 18), 
      axis.title.x=element_text(size=18),
      axis.title.y=element_text(size=18))+
  geom_smooth(method="lm", colour="red",linetype="dashed")+
  stat_regline_equation(label.x = -2.3, label.y = 4, size = 6,colour="grey30")+
        stat_cor( label.x=-2.3, label.y=3.3, size=6, colour="red")


```

```{r expression of genes of interest from DTU analysis}
genes_hgnc=c("RPL22L1", "GRB2", "RNF220")
genes=gsva_an$geneid[gsva_an$hgnc_symbol %in% genes_hgnc]
vsd_sel=as.data.frame(vsd_mat[rownames(vsd_mat) %in% genes,])
vsd_t=data.frame(t(vsd_sel))
vsd_t$Condition=sampleinfo$Condition
vsd_t$Experiment=sampleinfo$Experiment
colnames(vsd_t)=c(genes_hgnc, "Condition", "Experiment")

vsd_m=pivot_longer(vsd_t, cols=1:3)

#RPL22L1
ggplot(vsd_m[vsd_m$name=="RPL22L1",], aes(x=Condition, y=value, fill=Experiment))+ stat_summary(fun="mean", geom="bar") + stat_summary(geom="errorbar", colour="black", position=position_dodge(0.95))+ facet_wrap(~Experiment) +theme_bw()+ scale_fill_manual(values=c("turquoise3", "navy"))+xlab("Condition")+ylab("log2 Normalised  Expression") + theme(
      legend.text = element_text(size =14),
      axis.text = element_text(size = 18), 
      axis.title.x=element_text(size=18),
      axis.title.y=element_text(size=18))

#GRB2
ggplot(vsd_m[vsd_m$name=="GRB2",], aes(x=Condition, y=value, fill=Experiment))+ stat_summary(fun="mean", geom="bar") + stat_summary(geom="errorbar", colour="black", position=position_dodge(0.95))+ facet_wrap(~Experiment) +theme_bw()+ scale_fill_manual(values=c("turquoise3", "navy"))+xlab("Condition")+ylab("log2 Normalised  Expression") + theme(
      legend.text = element_text(size =14),
      axis.text = element_text(size = 18), 
      axis.title.x=element_text(size=18),
      axis.title.y=element_text(size=18))

#RNF220
ggplot(vsd_m[vsd_m$name=="RNF220",], aes(x=Condition, y=value, fill=Experiment))+ stat_summary(fun="mean", geom="bar") + stat_summary(geom="errorbar", colour="black", position=position_dodge(0.95))+ facet_wrap(~Experiment) +theme_bw()+ scale_fill_manual(values=c("turquoise3", "navy"))+xlab("Condition")+ylab("log2 Normalised  Expression") + theme(
      legend.text = element_text(size =14),
      axis.text = element_text(size = 18), 
      axis.title.x=element_text(size=18),
      axis.title.y=element_text(size=18))

#Pull out DGE significance values
cdna_dge[cdna_dge$hgnc %in% genes_hgnc,6:7]
drna_dge[drna_dge$hgnc %in% genes_hgnc, 6:7]
```


```{r}
---
title: "Gene Level Comparison"
author: "Rebecca L"
date: "2025-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggpubr)
library(ggpattern)
library(tidyverse)
library(ggplot2)
library(pheatmap)
library(DESeq2)
library(kableExtra)
library(vsn)
library(ggfortify)
library(AnnotationDbi)
library (org.Hs.eg.db)
library(PCAtools)
library(reshape2)
library(RColorBrewer)
library(fgsea)
library(GOSemSim)
library(enrichplot)
library(clusterProfiler)
library(ggvenn)
library(gplots)
library(purrr)
library(biomaRt)
library(GSEABase)
library(GSVA)
source("customfunc.R")

# Create PDF device for all plots
pdf("all_graphs_output.pdf", width = 12, height = 8)
```

```{r import, normalisation, PCA plot}
#Count matrix
cdna_count=read.table("cdna2_SKMM2_counts_gene.txt", header=T, sep="\t")[1:7]
drna_count=read.table("drnasub_SKMM2_counts_gene.txt", header=T, sep="\t")
combdata=left_join(cdna_count, drna_count, by="GENEID")

#Annotate HGNC symbols
gene_info=read.csv("tx_v46_info.csv", header=TRUE)[,c(2:3)]
readcounts_an=left_join(combdata, gene_info, by=join_by(GENEID==GENENAME), multiple="any")
rownames(readcounts_an)=readcounts_an$GENEID

#Sample Information
sampleinfo=read.table("sampleinfo.csv", header=T, row.names=1, sep=",", stringsAsFactors = TRUE)

#Lib size plot
librarySizes <- colSums(readcounts_an[,2:13])
barplot(librarySizes, names=names(librarySizes), las=2, main="Barplot of library sizes")

#Filter out 'IG' genes
ig=grep('^IGH|^IGK|^IGL|^HLA', readcounts_an$HGNC)
readcounts_an=readcounts_an[-ig,]

#Make DESeq Dataset & Filter No Counts
readcounts_an[,2:13]=sapply(readcounts_an[,2:13],as.integer)
dds=DESeqDataSetFromMatrix(countData=readcounts_an[,2:13],
                            colData=sampleinfo,
                            design=~Experiment+Condition)
hgnc=readcounts_an$HGNC
mcols(dds)= DataFrame(mcols(dds), hgnc)
mcols(dds)

#DESeq Object
dds <- estimateSizeFactors(dds)
dds=dds[rowSums(counts(dds))>=1,]

#DESeq + VSD Normalisation
dds=DESeq(dds)
vsd=vst(dds, blind=TRUE)
vsd_mat=as.matrix(assay(vsd))
meanSdPlot(vsd_mat, ranks=FALSE)

#PCA Plot
pca_res=prcomp(t(vsd_mat), scale. = TRUE)
print(autoplot(pca_res, data=sampleinfo, colour="Condition", shape="Experiment", size=4))

sampleDists <- dist(t(assay(vsd)))
library("RColorBrewer")
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$Experiment, vsd$Condition, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Reds")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```

```{r import DGE results}
#Read in results
cdna_dge=read.csv("DGE_cdna.csv", row.names=1, header=TRUE)
drna_dge=read.csv("DGE_drnasub.csv", row.names=1, header=TRUE)

#Remove NA
cdna_dge=cdna_dge[!is.na(cdna_dge$padj),]
drna_dge=drna_dge[!is.na(drna_dge$padj),]

#Extract sig up and sig down genes
cdna_up=rownames(cdna_dge)[cdna_dge$padj<0.05 & cdna_dge$log2FoldChange>0]
drna_up=rownames(drna_dge)[drna_dge$padj<0.05 & drna_dge$log2FoldChange>0]
up=list("PCR-cDNA"=cdna_up, directRNA=drna_up)
rm(cdna_up, drna_up)

cdna_down=rownames(cdna_dge)[cdna_dge$padj<0.05 & cdna_dge$log2FoldChange<0]
drna_down=rownames(drna_dge)[drna_dge$padj<0.05 & drna_dge$log2FoldChange<0]
down=list("PCR-cDNA"=cdna_down, directRNA=drna_down)
rm(cdna_down, drna_down)

#Venns
library(ggvenn)
#UP
print(ggvenn(up,
       fill_color=c("turquoise3", "navy"),
       text_size=6,
       show_percentage=FALSE,
        auto_scale=TRUE))

print(ggvenn(down,
        fill_color=c("turquoise3", "navy"),
       text_size=6,
       show_percentage=FALSE,
        auto_scale=TRUE))
```

```{r gsva}
#Get entrez IDs
ensembl=useEnsembl(biomart="genes",
                   dataset="hsapiens_gene_ensembl")
G_list=getBM(attributes=c("ensembl_gene_id_version", "entrezgene_id", "hgnc_symbol"), 
values=readcounts_an$GENEID, mart=ensembl)

gsva_array=as.data.frame(vsd_mat)
gsva_array$geneid=rownames(gsva_array)
gsva_an=inner_join(gsva_array, G_list, by=join_by(geneid==ensembl_gene_id_version), multiple="any")
gsva_an=gsva_an[!duplicated(gsva_an$hgnc_symbol),]
rownames(gsva_an)=gsva_an$hgnc_symbol

#Set up array
pdata=AnnotatedDataFrame(sampleinfo)
exprs=ExpressionSet(assayData=as.matrix(gsva_an[,1:12]), phenoData=pdata)

#Hallmark gene set
hallmark=list.files()
h.gs=getGmt("h.all.v2024.1.Hs.symbols.gmt", collectionType=BroadCollection(category="h"),
            geneIdType = SymbolIdentifier())
gsva_par=gsvaParam(exprs, h.gs, minSize=5, maxSize=500)
gsva=gsva(gsva_par)

#DE analysis of gene sets
library(limma)
library(edgeR)
gsva$group=paste0(gsva$Experiment,gsva$Condition)
mod=model.matrix(~gsva$group)
contrasts=makeContrasts("cDNAControl-cDNAIL6", "dRNAControl-dRNAIL6", 
levels=c("cDNAIL6", "cDNAControl", "dRNAIL6", "dRNAControl"))
fit=lmFit(gsva, mod)
fit=contrasts.fit(fit, contrasts)
fit=eBayes(fit)
res=decideTests(fit, p.value=0.05)
summary(res)

tt=topTable(fit, coef=1, n=50)
tt2=topTable(fit, coef=2, n=50)

#Format results df
pathways=rownames(tt)
pathways=strsplit(pathways, split="HALLMARK_" )
rownames(tt)=lapply(pathways, '[', 2)
tt=rownames_to_column(tt, var="pathway")
tt$cdna_sig=case_when(tt$adj.P.Val<0.05 & tt$logFC>0 ~'UP', 
tt$adj.P.Val<0.05 & tt$logFC<0 ~"DOWN", .default="NS")
tt=tt[,c(1,8)]

pathways=rownames(tt2)
pathways=strsplit(pathways, split="HALLMARK_")
rownames(tt2)=lapply(pathways, "[", 2)
tt2=rownames_to_column(tt2, var="pathway")
tt2$drna_sig=case_when(tt2$adj.P.Val<0.05 & tt2$logFC>0 ~'UP', 
tt2$adj.P.Val<0.05 & tt2$logFC<0 ~"DOWN", .default="NS")
tt2=tt2[, c(1,8)]

ann=inner_join(tt, tt2, by="pathway")
ann=column_to_rownames(ann, var="pathway")

#Annotate each gene set with up down or NS
ann_colour=list(cdna_sig=c(NS="grey30", UP="red", DOWN="mediumblue"),
                drna_sig=c(NS="grey30", UP="red", DOWN="mediumblue"),
                Condition=c(Control="pink", IL6="darkred"),
                Experiment=c(cDNA="turquoise3", dRNA="navy"))

pathways=rownames(gsva)
pathways=strsplit(pathways, split="HALLMARK_" )
gsva_array=exprs(gsva)
rownames(gsva_array)=lapply(pathways, '[', 2)

heatmap=pheatmap(gsva_array, clustering_distance_cols="correlation", 
clustering_distance_rows="correlation", annotation_colors=ann_colour, 
annotation_row=ann, annotation_col=sampleinfo, cutree_cols=2, cutree_rows=2, scale="column")
```

```{r scatterplots illustrative}
#Annotate protocol and bind
dge_comb=left_join(cdna_dge[,c(2,7)], drna_dge[,c(2,7)],by="hgnc")
colnames(dge_comb)=c("PCR-cDNA", "hgnc", "DirectRNA")
dge_comb_m=pivot_longer(dge_comb, cols=c(1,3))

scatter_plot <- ggplot(dge_comb, aes(x=`PCR-cDNA`, y=DirectRNA))+
geom_point(colour="grey30")+  
scale_y_continuous(limits=c(-2.5,5))+  
scale_x_continuous(limits=c(-2.5,5))+
theme(      
legend.text = element_text(size =14),      
axis.text = element_text(size = 18),       
axis.title.x=element_text(size=18),      
axis.title.y=element_text(size=18))+  
geom_smooth(method="lm", colour="red",linetype="dashed")+  
stat_regline_equation(label.x = -2.3, label.y = 4, size = 6,colour="grey30")+        
stat_cor( label.x=-2.3, label.y=3.3, size=6, colour="red")

print(scatter_plot)
```

```{r expression of genes of interest from DTU analysis}
genes_hgnc=c("RPL22L1", "GRB2", "RNF220")
genes=gsva_an$geneid[gsva_an$hgnc_symbol %in% genes_hgnc]
vsd_sel=as.data.frame(vsd_mat[rownames(vsd_mat) %in% genes,])
vsd_t=data.frame(t(vsd_sel))
vsd_t$Condition=sampleinfo$Condition
vsd_t$Experiment=sampleinfo$Experiment
colnames(vsd_t)=c(genes_hgnc, "Condition", "Experiment")
vsd_m=pivot_longer(vsd_t, cols=1:3)

#RPL22L1
rpl22l1_plot <- ggplot(vsd_m[vsd_m$name=="RPL22L1",], aes(x=Condition, y=value, fill=Experiment))+ 
stat_summary(fun="mean", geom="bar") + 
stat_summary(geom="errorbar", colour="black", position=position_dodge(0.95))+ 
facet_wrap(~Experiment) +
theme_bw()+ 
scale_fill_manual(values=c("turquoise3", "navy"))+
xlab("Condition")+
ylab("log2 Normalised  Expression") + 
theme(      
legend.text = element_text(size =14),      
axis.text = element_text(size = 18),       
axis.title.x=element_text(size=18),      
axis.title.y=element_text(size=18))

print(rpl22l1_plot)

#GRB2
grb2_plot <- ggplot(vsd_m[vsd_m$name=="GRB2",], aes(x=Condition, y=value, fill=Experiment))+ 
stat_summary(fun="mean", geom="bar") + 
stat_summary(geom="errorbar", colour="black", position=position_dodge(0.95))+ 
facet_wrap(~Experiment) +
theme_bw()+ 
scale_fill_manual(values=c("turquoise3", "navy"))+
xlab("Condition")+
ylab("log2 Normalised  Expression") + 
theme(      
legend.text = element_text(size =14),      
axis.text = element_text(size = 18),       
axis.title.x=element_text(size=18),      
axis.title.y=element_text(size=18))

print(grb2_plot)

#RNF220
rnf220_plot <- ggplot(vsd_m[vsd_m$name=="RNF220",], aes(x=Condition, y=value, fill=Experiment))+ 
stat_summary(fun="mean", geom="bar") + 
stat_summary(geom="errorbar", colour="black", position=position_dodge(0.95))+ 
facet_wrap(~Experiment) +
theme_bw()+ 
scale_fill_manual(values=c("turquoise3", "navy"))+
xlab("Condition")+
ylab("log2 Normalised  Expression") + 
theme(      
legend.text = element_text(size =14),      
axis.text = element_text(size = 18),       
axis.title.x=element_text(size=18),      
axis.title.y=element_text(size=18))

print(rnf220_plot)

#Pull out DGE significance values
cdna_dge[cdna_dge$hgnc %in% genes_hgnc,6:7]
drna_dge[drna_dge$hgnc %in% genes_hgnc, 6:7]

# Close the PDF device
dev.off()
```


---
title: "Gene Level Comparison"
author: "Rebecca L"
date: "2025-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggpubr)
library(ggpattern)
library(tidyverse)
library(ggplot2)
library(pheatmap)
library(DESeq2)
library(kableExtra)
library(vsn)
library(ggfortify)
library(AnnotationDbi)
library (org.Hs.eg.db)
library(PCAtools)
library(reshape2)
library(RColorBrewer)
library(fgsea)
library(GOSemSim)
library(enrichplot)
library(clusterProfiler)
library(ggvenn)
library(gplots)
library(purrr)
library(biomaRt)
library(GSEABase)
library(GSVA)
source("customfunc.R")
```

```{r import, normalisation, PCA plot}
#Count matrix
cdna_count=read.table("cdna2_SKMM2_counts_gene.txt", header=T, sep="\t")[1:7]
drna_count=read.table("drnasub_SKMM2_counts_gene.txt", header=T, sep="\t")
combdata=left_join(cdna_count, drna_count, by="GENEID")

#Annotate HGNC symbols
gene_info=read.csv("tx_v46_info.csv", header=TRUE)[,c(2:3)]
readcounts_an=left_join(combdata, gene_info, by=join_by(GENEID==GENENAME), multiple="any")
rownames(readcounts_an)=readcounts_an$GENEID

#Sample Information
sampleinfo=read.table("sampleinfo.csv", header=T, row.names=1, sep=",", stringsAsFactors = TRUE)

#Lib size plot
librarySizes <- colSums(readcounts_an[,2:13])
pdf("library_sizes_barplot.pdf", width = 10, height = 6)
barplot(librarySizes, names=names(librarySizes), las=2, main="Barplot of library sizes")
dev.off()

#Filter out 'IG' genes
ig=grep('^IGH|^IGK|^IGL|^HLA', readcounts_an$HGNC)
readcounts_an=readcounts_an[-ig,]

#Make DESeq Dataset & Filter No Counts
readcounts_an[,2:13]=sapply(readcounts_an[,2:13],as.integer)
dds=DESeqDataSetFromMatrix(countData=readcounts_an[,2:13],
                            colData=sampleinfo,
                            design=~Experiment+Condition)
hgnc=readcounts_an$HGNC
mcols(dds)= DataFrame(mcols(dds), hgnc)
mcols(dds)

#DESeq Object
dds <- estimateSizeFactors(dds)
dds=dds[rowSums(counts(dds))>=1,]

#DESeq + VSD Normalisation
dds=DESeq(dds)
vsd=vst(dds, blind=TRUE)
vsd_mat=as.matrix(assay(vsd))

# Save meanSdPlot
pdf("meanSd_plot.pdf", width = 8, height = 6)
meanSdPlot(vsd_mat, ranks=FALSE)
dev.off()

#PCA Plot
pca_res=prcomp(t(vsd_mat), scale. = TRUE)
pca_plot <- autoplot(pca_res, data=sampleinfo, colour="Condition", shape="Experiment", size=4)
ggsave("pca_plot.pdf", pca_plot, width = 10, height = 8)

sampleDists <- dist(t(assay(vsd)))
library("RColorBrewer")
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$Experiment, vsd$Condition, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Reds")) )(255)

# Save sample distance heatmap
pdf("sample_distance_heatmap.pdf", width = 10, height = 8)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
dev.off()
```

```{r import DGE results}
#Read in results
cdna_dge=read.csv("DGE_cdna.csv", row.names=1, header=TRUE)
drna_dge=read.csv("DGE_drnasub.csv", row.names=1, header=TRUE)

#Remove NA
cdna_dge=cdna_dge[!is.na(cdna_dge$padj),]
drna_dge=drna_dge[!is.na(drna_dge$padj),]

#Extract sig up and sig down genes
cdna_up=rownames(cdna_dge)[cdna_dge$padj<0.05 & cdna_dge$log2FoldChange>0]
drna_up=rownames(drna_dge)[drna_dge$padj<0.05 & drna_dge$log2FoldChange>0]
up=list("PCR-cDNA"=cdna_up, directRNA=drna_up)
rm(cdna_up, drna_up)

cdna_down=rownames(cdna_dge)[cdna_dge$padj<0.05 & cdna_dge$log2FoldChange<0]
drna_down=rownames(drna_dge)[drna_dge$padj<0.05 & drna_dge$log2FoldChange<0]
down=list("PCR-cDNA"=cdna_down, directRNA=drna_down)
rm(cdna_down, drna_down)

#Venns
library(ggvenn)
#UP
up_venn <- ggvenn(up,
       fill_color=c("turquoise3", "navy"),
       text_size=6,
       show_percentage=FALSE,
        auto_scale=TRUE)
ggsave("venn_upregulated_genes.pdf", up_venn, width = 8, height = 6)

down_venn <- ggvenn(down,
        fill_color=c("turquoise3", "navy"),
       text_size=6,
       show_percentage=FALSE,
        auto_scale=TRUE)
ggsave("venn_downregulated_genes.pdf", down_venn, width = 8, height = 6)
```

```{r gsva}
#Get entrez IDs
ensembl=useEnsembl(biomart="genes",
                   dataset="hsapiens_gene_ensembl")
G_list=getBM(attributes=c("ensembl_gene_id_version", "entrezgene_id", "hgnc_symbol"), 
values=readcounts_an$GENEID, mart=ensembl)

gsva_array=as.data.frame(vsd_mat)
gsva_array$geneid=rownames(gsva_array)
gsva_an=inner_join(gsva_array, G_list, by=join_by(geneid==ensembl_gene_id_version), multiple="any")
gsva_an=gsva_an[!duplicated(gsva_an$hgnc_symbol),]
rownames(gsva_an)=gsva_an$hgnc_symbol

#Set up array
pdata=AnnotatedDataFrame(sampleinfo)
exprs=ExpressionSet(assayData=as.matrix(gsva_an[,1:12]), phenoData=pdata)

#Hallmark gene set
hallmark=list.files()
h.gs=getGmt("h.all.v2024.1.Hs.symbols.gmt", collectionType=BroadCollection(category="h"),
            geneIdType = SymbolIdentifier())
gsva_par=gsvaParam(exprs, h.gs, minSize=5, maxSize=500)
gsva=gsva(gsva_par)

#DE analysis of gene sets
library(limma)
library(edgeR)
gsva$group=paste0(gsva$Experiment,gsva$Condition)
mod=model.matrix(~gsva$group)
contrasts=makeContrasts("cDNAControl-cDNAIL6", "dRNAControl-dRNAIL6", 
levels=c("cDNAIL6", "cDNAControl", "dRNAIL6", "dRNAControl"))
fit=lmFit(gsva, mod)
fit=contrasts.fit(fit, contrasts)
fit=eBayes(fit)
res=decideTests(fit, p.value=0.05)
summary(res)

tt=topTable(fit, coef=1, n=50)
tt2=topTable(fit, coef=2, n=50)

#Format results df
pathways=rownames(tt)
pathways=strsplit(pathways, split="HALLMARK_" )
rownames(tt)=lapply(pathways, '[', 2)
tt=rownames_to_column(tt, var="pathway")
tt$cdna_sig=case_when(tt$adj.P.Val<0.05 & tt$logFC>0 ~'UP', 
tt$adj.P.Val<0.05 & tt$logFC<0 ~"DOWN", .default="NS")
tt=tt[,c(1,8)]

pathways=rownames(tt2)
pathways=strsplit(pathways, split="HALLMARK_")
rownames(tt2)=lapply(pathways, "[", 2)
tt2=rownames_to_column(tt2, var="pathway")
tt2$drna_sig=case_when(tt2$adj.P.Val<0.05 & tt2$logFC>0 ~'UP', 
tt2$adj.P.Val<0.05 & tt2$logFC<0 ~"DOWN", .default="NS")
tt2=tt2[, c(1,8)]

ann=inner_join(tt, tt2, by="pathway")
ann=column_to_rownames(ann, var="pathway")

#Annotate each gene set with up down or NS
ann_colour=list(cdna_sig=c(NS="grey30", UP="red", DOWN="mediumblue"),
                drna_sig=c(NS="grey30", UP="red", DOWN="mediumblue"),
                Condition=c(Control="pink", IL6="darkred"),
                Experiment=c(cDNA="turquoise3", dRNA="navy"))

pathways=rownames(gsva)
pathways=strsplit(pathways, split="HALLMARK_" )
gsva_array=exprs(gsva)
rownames(gsva_array)=lapply(pathways, '[', 2)

# Save GSVA heatmap
pdf("gsva_hallmark_heatmap.pdf", width = 8, height = 10)
heatmap=pheatmap(gsva_array, clustering_distance_cols="correlation", 
clustering_distance_rows="correlation", annotation_colors=ann_colour, 
annotation_row=ann, annotation_col=sampleinfo, cutree_cols=2, cutree_rows=2, scale="column")
dev.off()
```

```{r scatterplots illustrative}
#Annotate protocol and bind
dge_comb=left_join(cdna_dge[,c(2,7)], drna_dge[,c(2,7)],by="hgnc")
colnames(dge_comb)=c("PCR-cDNA", "hgnc", "DirectRNA")
dge_comb_m=pivot_longer(dge_comb, cols=c(1,3))

scatter_plot <- ggplot(dge_comb, aes(x=`PCR-cDNA`, y=DirectRNA))+
  geom_point(size = 0.8, alpha = 0.6, color = "black") +  
  scale_y_continuous(limits = c(-2.5, 5), 
                     breaks = seq(-2, 5, 1),
                     expand = c(0.02, 0)) +  
  scale_x_continuous(limits = c(-2.5, 5), 
                     breaks = seq(-2, 5, 1),
                     expand = c(0.02, 0)) +
  geom_smooth(method = "lm", color = "red", linewidth = 0.8, se = FALSE) +
  labs(x = "PCR-cDNA log₂FC", 
       y = "Direct RNA log₂FC") +
  stat_regline_equation(label.x = -2.2, label.y = 4.5, 
                       size = 2.5, color = "black") +        
  stat_cor(label.x = -2.2, label.y = 4.0, 
           size = 2.5, color = "black") +
  theme_classic() +
  theme(
    axis.text = element_text(size = 8, color = "black"),       
    axis.title = element_text(size = 9, color = "black"),
    axis.line = element_line(color = "black", linewidth = 0.5),
    axis.ticks = element_line(color = "black", linewidth = 0.5),
    axis.ticks.length = unit(2, "pt"),
    plot.margin = margin(10, 10, 10, 10),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA)
  ) +
  coord_fixed(ratio = 1)

ggsave("dge_correlation_scatterplot.pdf", scatter_plot, 
       width = 3.5, height = 3.5, units = "in", dpi = 300)
```


```{r expression of genes of interest from DTU analysis}
genes_hgnc=c("RPL22L1", "GRB2", "RNF220")
genes=gsva_an$geneid[gsva_an$hgnc_symbol %in% genes_hgnc]
vsd_sel=as.data.frame(vsd_mat[rownames(vsd_mat) %in% genes,])
vsd_t=data.frame(t(vsd_sel))
vsd_t$Condition=sampleinfo$Condition
vsd_t$Experiment=sampleinfo$Experiment
colnames(vsd_t)=c(genes_hgnc, "Condition", "Experiment")
vsd_m=pivot_longer(vsd_t, cols=1:3)

#RPL22L1
rpl22l1_plot <- ggplot(vsd_m[vsd_m$name=="RPL22L1",], aes(x=Condition, y=value, fill=Experiment))+ 
stat_summary(fun="mean", geom="bar") + 
stat_summary(geom="errorbar", colour="black", position=position_dodge(0.95))+ 
facet_wrap(~Experiment) +
theme_bw()+ 
scale_fill_manual(values=c("turquoise3", "navy"))+
xlab("Condition")+
ylab("log2 Normalised  Expression") + 
theme(      
legend.text = element_text(size =14),      
axis.text = element_text(size = 18),       
axis.title.x=element_text(size=18),      
axis.title.y=element_text(size=18))

ggsave("RPL22L1_expression.pdf", rpl22l1_plot, width = 10, height = 6)

#GRB2
grb2_plot <- ggplot(vsd_m[vsd_m$name=="GRB2",], aes(x=Condition, y=value, fill=Experiment))+ 
stat_summary(fun="mean", geom="bar") + 
stat_summary(geom="errorbar", colour="black", position=position_dodge(0.95))+ 
facet_wrap(~Experiment) +
theme_bw()+ 
scale_fill_manual(values=c("turquoise3", "navy"))+
xlab("Condition")+
ylab("log2 Normalised  Expression") + 
theme(      
legend.text = element_text(size =14),      
axis.text = element_text(size = 18),       
axis.title.x=element_text(size=18),      
axis.title.y=element_text(size=18))

ggsave("GRB2_expression.pdf", grb2_plot, width = 10, height = 6)

#RNF220
rnf220_plot <- ggplot(vsd_m[vsd_m$name=="RNF220",], aes(x=Condition, y=value, fill=Experiment))+ 
stat_summary(fun="mean", geom="bar") + 
stat_summary(geom="errorbar", colour="black", position=position_dodge(0.95))+ 
facet_wrap(~Experiment) +
theme_bw()+ 
scale_fill_manual(values=c("turquoise3", "navy"))+
xlab("Condition")+
ylab("log2 Normalised  Expression") + 
theme(      
legend.text = element_text(size =14),      
axis.text = element_text(size = 18),       
axis.title.x=element_text(size=18),      
axis.title.y=element_text(size=18))

ggsave("RNF220_expression.pdf", rnf220_plot, width = 10, height = 6)

#Pull out DGE significance values
cdna_dge[cdna_dge$hgnc %in% genes_hgnc,6:7]
drna_dge[drna_dge$hgnc %in% genes_hgnc, 6:7]
```